Synthesis of Linear Functional Programs

### 2021-07-11 article/article.tex
Write architecture and implementation start

% General motivation?
Program synthesis is an automated or semi-automated process of
deriving a program (i.e.~generating code) from a high-level
specification.  
%
Synthesis can be seen as a means to improve a programmer's
productivity and/or program correctness (i.e. through suggestion
and/or autocompletion), or as a tool to automate certain parts of the
programming process (e.g. in the same way AI might generate some
boilerplate text to aid a journalist, rather than to replace the
journalist outright), just to name a few examples.

Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind
of specification, program synthesizers have to deal with two main
inherent sources of complexity: searching over a vast space of
(potentially) valid programs, and interpreting user intent.
%
In this work we will explore type-based synthesis where specifications
take the form of linear types, which are types that limit resource
usage in programs.  Concretely, we will explore the proof-theoretic
foundations of linear types via the Curry-Howard correspondence with
linear logic, viewing program synthesis under the lens of (linear
logic) bottom-up proof search.

@@

Richer type systems allow for more precise types, which can
statically eliminate various kinds of logical errors by making certain
invalid program states ill-typed (e.g.,~a ``null aware'' type system,
will ensure at compile-time that you cannot dereference a
null-pointer). However, they can also be a burden -- the whole point
of these type systems is to ensure that \emph{fewer} programs are
deemed well-typed, which can pose additional challenges to the
development process. Type-based synthesis leverages rich types as a
way of pruning the search space, and by using types the user 
is given a more ``familiar'' specification. For instance, the type
$\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
specifies a (curried) function that takes two integers and produces an
integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).
However, the richer type $(x{:}\mathsf{Int}) \rightarrow
(y{:}\mathsf{Int}) \rightarrow \{z{:}\mathsf{Int} \mid z = x+y\}$
very precisely specifies a function that takes integer arguments $x$
and $y$ and returns an integer $z$ that is the sum of $x$ and $y$.

%
This work explores the synthesis of linear functional
programs from types based on linear logic propositions (i.e.~linear
types) by leveraging the Curry-Howard correspondence. It
states that propositions in a logic have a direct
mapping to types, and well-typed programs correspond to proofs of
those propositions.  As such, having a type be a proposition in linear
logic, we can relate a proof of it directly to a linear
functional program — finding a proof is finding a program with that
type. Formulating synthesis as proof search in linear logic
allows us to inform our work with approaches from the relevant 
literature.

@@

\mypara{Resource-Management}\label{sec:background_resource_management}
Be it from the perspective of proof \emph{checking} (i.e.~type checking) or
proof \emph{search} linear logic poses a key challenge when compared
to the non-linear setting:
When constructing a derivation (bottom-up), we are seemingly forced to
guess how to correctly split the linear context such that the
sub-derivations have access to the correct resources (e.g.~the
$\tensor I$ rule above).
%
%

\mypara{Focusing}\label{sec:background_focusing}

@@

Escrever introdução melhor que tenha isto como seguimento...

At the core of the synthesis process, and of the developed synthetiser \synname,
is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
resource management. It connects the concepts seen above -- a type system
consisting of inference rules in sequent calculus for linear logic connectives;
with resource management, and focusing, as formalized techniques to,
respectively, solve resource related non-determinism, and to eliminate from the
proof search don't-care non-determinism.  Guided solely by this system's formal
representation\ref{sec:formal_system}, we can (and did) create a very robust
synthetiser for a select part of the linear types that's able to synth
programs successfully from e.g. the major linear logic equivalences. For example,
given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
the program $\lambda a \rightarrow \lambda b \rightarrow$ let $c \tensor d$ =
$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.

@@

\subsection{The \synname\ language}

@@

\mypara{Core Rules} The core system comprises of the sequent calculus rules for
linear logic connectives with resource-management and focusing. As seen
previously\ref{sec:background}, focusing separates the proof search into two
phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later try to
invert, moving them to the linear context ($\Delta$) when we fail to.

@@

@@

>>>>

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program (i.e.~generating code) from a high-level specification.
%
Synthesis can be seen as a means to improve a programmer's productivity and/or
program correctness (i.e. through suggestion and/or autocompletion), or as a
tool to automate certain parts of the programming process, to name a few
examples.

Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind of
specification, program synthesizers have to deal with two main inherent sources
of complexity: searching over a vast space of (potentially) valid programs, and
interpreting user intent.
%
In this work we'll explore type-based synthesis where specifications take the
form of linear types, which limit resource usage in programs, in combination
with polymorphism.  Concretely, we will explore the proof-theoretic foundations
of linear types via the Curry-Howard correspondence with linear logic, viewing
program synthesis under the lens of (linear logic) bottom-up proof search.

@@

Richer type systems allow for more precise types, which can
statically eliminate various kinds of logical errors by making certain
invalid program states ill-typed (e.g.,~a ``null aware'' type system,
will ensure at compile-time that you cannot dereference a
null-pointer). However, they can also be a burden -- the whole point
of these type systems is to ensure that \emph{fewer} programs are
deemed well-typed, which can pose additional challenges to the
development process. Type-based synthesis leverages rich types as a
way of pruning the search space, and by using types the user 
is given a more ``familiar'' specification interface. For instance, the type
$\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
specifies a (curried) function that takes two integers and produces an
integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).
However, the richer type $(x{:}\mathsf{Int}) \rightarrow
(y{:}\mathsf{Int}) \rightarrow \{z{:}\mathsf{Int} \mid z = x+y\}$
very precisely specifies a function that takes integer arguments $x$
and $y$ and returns an integer $z$ that is the sum of $x$ and $y$.

%
This work explores the synthesis of linear functional
programs from types based on linear logic propositions (i.e.~linear
types) by leveraging the Curry-Howard correspondence.

% It
% states that propositions in a logic have a direct
% mapping to types, and well-typed programs correspond to proofs of
% those propositions.  As such, having a type be a proposition in linear
% logic, we can relate a proof of it directly to a linear
% functional program — finding a proof is finding a program with that
% type. Formulating synthesis as proof search in linear logic
% allows us to inform our work with approaches from the relevant 
% literature.

@@

@@

Escrever introdução melhor que tenha isto como seguimento...e que siga duma
introdução extendida

At the core of the synthesis process, and of the developed synthetiser \synname,
is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
resource management. It connects the concepts seen in
background~\ref{sec:background} -- a type system
consisting of inference rules in sequent calculus for linear logic connectives;
with resource management, and focusing, as formalized techniques to,
respectively, solve resource related non-determinism, and to eliminate from the
proof search don't-care non-determinism.  Guided solely by this system's formal
representation\ref{sec:formal_system}, we can (and did) create a very robust
synthetiser for a select part of the linear types that's able to synth
programs successfully from e.g. the major linear logic equivalences. For example,
given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
the program $\lambda a \rightarrow \lambda b \rightarrow$ let $c \tensor d$ =
$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.

@@

@@

\mypara{Core Rules} The core system comprises of the sequent calculus rules for
linear logic connectives with resource-management and focusing. As explained in\ref{sec:background}, focusing separates the proof search into two
phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later try to
invert, moving them to the linear context ($\Delta$) when we fail to.

@@

\mypara{Polymorfic ADTs}

\mypara{Refinement Types}

The \synname\ synthetiser operates as part of the pipeline that processes a
full program in the \synname\ language, and is called to generate parts marked
(by the syntax \emph{\{\{ ... \}\}}) for
synthesis of said program. The main pipeline consists of:
\[
    \textrm{Parsing} \rightarrow \textrm{Desugaring} \rightarrow \textrm{Inference} \rightarrow \textrm{Synthesis}
\]

The parsing module converts ADT declarations and top level functions written in
the language's syntax to a list of abstract syntax trees (ASTs) understood by the rest of the
pipeline.

The desugaring module converts the frontend AST to a core AST without syntatic
sugar and using the locally-nameless~\cite{locally nameless} representation for
all bound variables. This way inference can be done without worrying about name
conflicts.

The inference module will traverse the list of ASTs in order (top-down), and
infer the type of all functions defined by the user, checking the inferred
type against a possible user given type annotation. Whenever a synth mark is
found in the AST, all the inferrence context up to that point is added to the
mark, and if not explicit, its type is inferred, essentially defining the synth
context and synth goal.

The synthesis module iterates over the synth marks and runs the synthetiser with
its context and goal for each, minimizing the resulting expression, and
replacing the mark in the frontend AST with the synthed program.

\mypara{General} The complete implementation of the synthetiser and environment
pipeline was done in Haskell...

\mypara{Backtracking} Even though the synthetiser formal system helps, there's
still non-determinism in the synth process. At diverse times in the proof
construction we must choose one of multiple rules to apply, and in the event of
not being able to construct the proof following that choice, we want to
backtrack to the decision point and attempt to create a proof taking a different
route. To this effect, our prototype synthetiser depends on \emph{logict}, "a backtracking logic-programming
monad"~\cite{logict}.

\mypara{SMT Solving} To typecheck and synth refinement types and dependent
functions, we need to contact an SMT solver to check satisfiability. For this,
we use the library \emph{sbv}~\cite{sbv}. Unfortunately, this library does not
support the usage of the SMT solvers to satisfy and output more complicated
first-order logic solutions -- since this is needed to synth refinement types
when right focused, we instead use unsafe library primitives to communicate with
the solver directly, constructing the logical formulas ourselves. For 
satisfying formulas, for instance, when typechecking, or left focused on a
refinement type, we use the library api successfully to get a veridct on
satisfiability.

\mypara{Interface} A simple, syntax-highlighted, web interface was developed
alongside a server to make possible experimenting with the \synname\ 
language without having to download and compile the complete toolchain. The web
interface is available from the repository~\cite{github repo}.

@@

\mypara{Resource-Management}
Be it from the perspective of proof \emph{checking} (i.e.~type checking) or
proof \emph{search} linear logic poses a key challenge when compared
to the non-linear setting:
When constructing a derivation (bottom-up), we are seemingly forced to
guess how to correctly split the linear context such that the
sub-derivations have access to the correct resources (e.g.~the
$\tensor I$ rule above).
%
%

\mypara{Focusing}
